---
title: "Exploration of real bulk RNA-seq datasets"
subtitle: "TB datasets"
author: Yuqing Zhang
date: "`r Sys.Date()`"
output:
  html_notebook:
    theme: united
    toc: yes
bibliography: refs.bib
---

## Introduction

This is a curated dataset containing the Africa, Brazil, India, and the G6 datasets, which are all TB datasets with active TB patients, patients that will progress to active TB, and people with TB infection who don’t progress. There is a batch effect especially with the Brazil data. This object is a SummarizedExperiment with 7 assays. The ‘combat’ assay was made by running the logtpm values through combat with ‘SequencingBatch’ as the batch and ‘Label’ as the covariate. “Label” is the condition Non-progressor, Progressor, and Active.


### Overall design

The combined dataset contains 4 studies, in which Brazil dataset has 2 batches within the study:

+ Africa: 197 samples, (104 Non-progressor, 77 Progressor, 16 Active)
+ Brazil: 51 samples, (21 Non-progressor, 16 Progressor, 14 Active)
    - Batch 1: 39 samples, (9 Non-progressor, 16 Progressor, 14 Active)
    - Batch 2: 12 samples, (12 Non-progressor, 0 Progressor, 0 Active)
+ G6: 399 samples, (304 Non-progressor, 95 Progressor, 0 Active)
+ India: 44 samples, (19 Non-progressor, 0 Progressor, 25 Active)

**Note**: Gender is an important factor in this study.


### Load data & summarize basic info

```{r, echo=FALSE, results='hide'}
rm(list=ls())
output_dir <- "~/Google Drive/ComBat_seq/real_data_example/RNAseq/TB"
sapply(c("SummarizedExperiment", "DESeq2", "edgeR", "dendextend", "ggplot2", "reshape2", 
         "gridExtra", "scales", "ggdendro", "MASS"), require, character.only=TRUE)

####  Load data
rds_obj <- readRDS(file.path(output_dir, "combined.rds"))
cts <- assays(rds_obj)$counts
# batch indicator
batch <- colData(rds_obj)$SequencingBatch
# condition indicator
group <- colData(rds_obj)$Label
```

After loading the data, we calculate a few basic statistics of the dataset.

```{r, echo=FALSE}
cat("######  Samples  #######\n")
cat("- Total number of samples:\n")
print(ncol(cts))

cat("\n- Number of samples in condition groups:\n")
print(table(group))

cat("\n- Number of samples in batches:\n")
print(table(batch))
cat("\nIn each batch separately:\n")
n_batches <- lapply(levels(batch), function(curr_batch){
  table(group[batch==curr_batch])
})
n_batches <- do.call(rbind, n_batches)
rownames(n_batches) <- levels(batch)
print(n_batches)


cat("\n\n######  Genes  #######\n")
cat("- Total number of genes:\n")
print(nrow(cts))

cat("\n- Number of genes with 0 counts:\n")
gene_row_with_zeros <- apply(cts, 1, function(x){any(x==0)})
cat(c(sum(gene_row_with_zeros), percent(sum(gene_row_with_zeros)/nrow(cts))))

cat("\n\n- Number of genes with maximum 10 counts:\n")
gene_row_low <- apply(cts, 1, function(x){all(x<=10)})
cat(c(sum(gene_row_low), percent(sum(gene_row_low)/nrow(cts))))

cat("\n\n- Number of genes with ONLY 0s:\n")
gene_all_zeros <- apply(cts, 1, function(x){all(x==0)})
cat(c(sum(gene_all_zeros), percent(sum(gene_all_zeros)/nrow(cts))))


cat("\n\n\n######  Counts  #######\n")
cat("- Range of count in whole dataset:\n")
print(round(range(cts), 3))

cat("\n- Average count in whole dataset:\n")
print(paste("Mean:", round(mean(cts), 3)))
print(paste("Median:", round(median(cts), 3)))

rm(gene_row_with_zeros, gene_row_low, gene_all_zeros)
```

There are no genes with only zero values across all samples (probably because of cleaning in the curation stage). The figure below shows distribution of percentages of zero counts for the remaining genes (each data point represents a gene, i.e., how many zeros are there in each gene).

```{r, echo=FALSE}
n_zeros_in_genes <- apply(cts, 1, function(x){length(which(x==0))})
percent_zeros_in_genes <- n_zeros_in_genes / ncol(cts)
percent_zeros_df <- data.frame(genes=rownames(cts), value=percent_zeros_in_genes)

binned_cts <- hist(percent_zeros_in_genes, breaks=10, plot=FALSE)$counts 
binned_proportions <- binned_cts / sum(binned_cts)
names(binned_proportions) <- paste0("<", percent(seq(from=0.1, to=1, by=0.1)))  
cat("######  Proportion of genes with X% zeros across samples  #######\n")
print(round(binned_proportions,3))

ggplot(percent_zeros_df, aes(0, value)) + 
      geom_violin() +
      geom_boxplot(aes(0.7, value), width=0.15) +
      coord_flip() +
      annotate(geom="text", label=percent(median(percent_zeros_df$value)), x=0.85, y=median(percent_zeros_df$value)) +
      labs(y="% Zeros", title="Proportion of zeros across samples for the remaining genes") +
      theme(axis.title.y=element_blank())#, axis.text.y=element_blank())
```

```{r, echo=FALSE}
rm(n_zeros_in_genes, percent_zeros_in_genes, percent_zeros_df)
```



## Existence of batch effect {.tabset}

In this merged dataset, there are both "study effects" across the 4 studies (targeting different populations), and batch effect within the same study (Brazil dataset).

### PCA

```{r, echo=FALSE, fig.height=3.5, fig.width=5}
pca_obj <- plotPCA(DESeqTransform(rds_obj), intgroup=c("SequencingBatch", "Label", "Sex", "Dataset")) 
p_batch <- ggplot(pca_obj$data, aes(x=PC1, y=PC2, color=SequencingBatch)) +
  geom_point() + 
  labs(x=sprintf("PC1: %s Variance", percent(pca_obj$plot_env$percentVar[1])),
       y=sprintf("PC2: %s Variance", percent(pca_obj$plot_env$percentVar[2])),
       title="PCA: samples colored by Batch")
p_cond <- ggplot(pca_obj$data, aes(x=PC1, y=PC2, color=Label)) +
  geom_point() + 
  labs(x=sprintf("PC1: %s Variance", percent(pca_obj$plot_env$percentVar[1])),
       y=sprintf("PC2: %s Variance", percent(pca_obj$plot_env$percentVar[2])),
       title="PCA: samples colored by Condition")
p_studies <- ggplot(pca_obj$data, aes(x=PC1, y=PC2, color=Dataset)) +
  geom_point() + 
  labs(x=sprintf("PC1: %s Variance", percent(pca_obj$plot_env$percentVar[1])),
       y=sprintf("PC2: %s Variance", percent(pca_obj$plot_env$percentVar[2])),
       title="PCA: samples colored by Studies")
p_sex <- ggplot(pca_obj$data, aes(x=PC1, y=PC2, color=Sex)) +
  geom_point() + 
  labs(x=sprintf("PC1: %s Variance", percent(pca_obj$plot_env$percentVar[1])),
       y=sprintf("PC2: %s Variance", percent(pca_obj$plot_env$percentVar[2])),
       title="PCA: samples colored by Sex")
grid.arrange(p_batch, p_cond, p_studies, p_sex, nrow=2, ncol=2)
```


### PCA (Brazil only)

```{r, echo=FALSE, fig.height=2, fig.width=5}
rds_brazil <- rds_obj[, rds_obj$Dataset=="Brazil"]

pca_obj_brazil <- plotPCA(DESeqTransform(rds_brazil), intgroup=c("SequencingBatch", "Label")) 
pb_batch <- ggplot(pca_obj_brazil$data, aes(x=PC1, y=PC2, color=SequencingBatch)) +
  geom_point() + 
  labs(x=sprintf("PC1: %s Variance", percent(pca_obj_brazil$plot_env$percentVar[1])),
       y=sprintf("PC2: %s Variance", percent(pca_obj_brazil$plot_env$percentVar[2])),
       title="PCA: samples colored by Batch")
pb_cond <- ggplot(pca_obj_brazil$data, aes(x=PC1, y=PC2, color=Label)) +
  geom_point() + 
  labs(x=sprintf("PC1: %s Variance", percent(pca_obj_brazil$plot_env$percentVar[1])),
       y=sprintf("PC2: %s Variance", percent(pca_obj_brazil$plot_env$percentVar[2])),
       title="PCA: samples colored by Condition")
grid.arrange(pb_batch, pb_cond, ncol=2)
```


### Dendrograms (Brazil only)

```{r, echo=FALSE, fig.height=2.8, fig.width=5.5}
counts_norm <- scale(t(assay(rds_brazil,"counts")), center=TRUE, scale=TRUE)
hc <- hclust(dist(counts_norm))
dend <- as.dendrogram(hc)
ddata <- dendro_data(dend, type="rectangle")

meta_info <- data.frame(colData(rds_brazil)[, c("SequencingBatch", "Label")], sample=colnames(rds_brazil))
tmp <- merge(x=ddata$labels, y=meta_info, by.x="label", by.y="sample")

pd_batch <- ggplot(segment(ddata)) + 
  geom_segment(aes(x=x, y=y, xend=xend, yend=yend)) + 
  coord_flip() + scale_y_reverse() +
  scale_color_manual(values=c('#E69F00', '#56B4E9')) +
  geom_text(data=tmp, aes(label=tmp$label, x=x, y=-10, colour=SequencingBatch)) +
  labs(x="", y="", title="Clustering: samples colored by Batch")
pd_cond <- ggplot(segment(ddata)) + 
  geom_segment(aes(x=x, y=y, xend=xend, yend=yend)) + 
  coord_flip() + scale_y_reverse() +
  geom_text(data=tmp, aes(label=tmp$label, x=x, y=-10, colour=Label)) +
  labs(x="", y="", title="Clustering: samples colored by Condition")
grid.arrange(pd_batch, pd_cond, ncol=2)

rm(counts_norm, ddata, dend, hc, p_batch, p_cond, p_studies, p_sex, pca_obj, pd_batch, pd_cond, tmp,
   pca_obj_brazil, pb_batch, pb_cond)
```



## Library size

Next, we take a look at the observed library sizes of samples, and to see if there is a difference in library size across batches.

```{r, echo=FALSE}
###  Distribution of library size
lib_sizes <- colSums(cts)
cat("Range of observed library sizes across samples:\n")
print(scientific(range(lib_sizes)))
```

```{r, echo=FALSE, fig.height=2, fig.width=4.5}
lib_sizes_df <- data.frame(lib_sizes=lib_sizes, batch=batch)

# overall sample library sizes
plib1 <- ggplot(lib_sizes_df, aes(x=lib_sizes)) + 
  geom_histogram(bins=round(ncol(rds_obj)/5)) +
  geom_vline(aes(xintercept=mean(lib_sizes)), color="red", linetype="dashed", size=1) +
  annotate(geom="text", x=mean(lib_sizes), y=max(hist(lib_sizes, plot=FALSE)$counts)+1, 
           label=scientific(mean(lib_sizes)), color="red") +
  labs(x="Library sizes in samples", y="Number of samples", 
       title="Distribution of observed library sizes") 

# boxplot comparing distribution of library size across batches
plib2 <- ggplot(lib_sizes_df, aes(x=batch, y=lib_sizes, fill=batch)) + 
  geom_boxplot() +
  labs(x="", y="Library sizes", title="Compare library sizes across batch") 

grid.arrange(plib1, plib2, ncol=2)

rm(plib1, plib2)
```



## Gene expression distribution

Our next step is to interrogate the gene expression distributions. 

<!--
We first simply plot the counts, CPM, and log2CPM of 20 randomly chosen genes in the dataset. These figures are stored under directory specified by {output_dir}/randGplots/.  
--> 

```{r, echo=FALSE, results='hide', eval=FALSE}
if(!dir.exists(file.path(output_dir, "randGplots"))){dir.create(file.path(output_dir, "randGplots"))}
set.seed(123)
N_sel <- 20
chosen_id <- sample(1:nrow(rds_obj), N_sel, replace=FALSE)

cts_subset <- cts[chosen_id, ]
cpm_subset <- cpm(cts, log=FALSE)[chosen_id, ]
logCPM_subset <- cpm(cts, log=TRUE)[chosen_id, ]
data_lst <- list(Counts=cts_subset, CPM=cpm_subset, log2CPM=logCPM_subset)

void <- mapply(function(dat, dat_name, meta_info){
  for(i in 1:nrow(dat)){
    # prepare data frame
    curr_df <- data.frame(samples=colnames(dat), value=dat[i, ])
    curr_df <- merge(x=curr_df, y=meta_info, by.x="samples", by.y="sample")
    
    # scatter plot for i-th gene
    p_scatter <- ggplot(curr_df, aes(x=samples, y=value, color=batch, shape=cond)) +
      geom_point() +
      labs(x="Samples", y=dat_name, title=rownames(dat)[i]) +
      theme(axis.text.x=element_blank())
    
    # violin plot for i-th gene
    p_violin <- ggplot(curr_df, aes(0, value)) + 
      geom_violin() +
      geom_boxplot(width=0.2) +
      coord_flip() +
      geom_text(aes(label=as.integer(median(curr_df$value)), x=0.12, y=median(curr_df$value))) +
      labs(y=dat_name, title=sprintf("All samples, %s", rownames(dat)[i])) +
      theme(axis.title.y=element_blank())
    
    # violin plot group by batch
    medians_batch <- aggregate(value ~ batch, data=curr_df, FUN=median)
    p_violin_batch <- ggplot(curr_df, aes(x=batch, y=value, color=batch)) + 
      geom_violin() +
      geom_boxplot(width=0.2) +
      coord_flip() +
      scale_color_manual(values=c('#E69F00', '#56B4E9')) +
      geom_text(data=medians_batch, 
                aes(label=round(medians_batch$value,2), x=as.numeric(as.character(medians_batch$batch))+0.12,
                    y=as.numeric(as.character(medians_batch$value)))) +
      labs(y=dat_name, title=sprintf("In batches, %s", rownames(dat)[i])) +
      theme(axis.title.y=element_blank())
    
    # violin plot group by condition
    medians_cond <- aggregate(value ~ cond, data=curr_df, FUN=median)
    p_violin_cond <- ggplot(curr_df, aes(x=cond, y=value, color=cond)) + 
      geom_violin() +
      geom_boxplot(width=0.2) +
      coord_flip() +
      geom_text(data=medians_cond,
                aes(label=round(medians_cond$value,2), x=as.numeric(as.character(medians_cond$cond))+1.12, 
                    y=as.numeric(as.character(medians_cond$value)))) +
      labs(y=dat_name, title=sprintf("In condition groups, %s", rownames(dat)[i])) +
      theme(axis.title.y=element_blank())
    
    # output plot
    png(sprintf("%s/%s_gene%s.png", file.path(output_dir, "randGplots"), dat_name, i),
        width=10, height=8, units="in", res=300)
    grid.arrange(p_scatter, p_violin, p_violin_batch, p_violin_cond, nrow=2, ncol=2)
    dev.off()
  }
}, dat=data_lst, dat_name=names(data_lst), meta_info=list(meta_info,meta_info,meta_info))

rm(cts_subset, cpm_subset, logCPM_subset, lib_sizes, lib_sizes_df, chosen_id, data_lst, void, N_sel)
```


### Fold changes in mean {.tabset}

We use edgeR to perform differential expression analysis, and calculate (log-) fold changes of all genes between condition groups and batches.

#### Change WRT condition

To rule out the impact of batch effect, we limit our DE analysis with regard to biological condition in one of the batches. Here we specify the batch to be **G6** due to a larger batch size, and evaluate DE between progressors and non-progressors.

```{r, results="hide"}
####  Differential genes relative to biological condition 
####  (within batch 1 / batch specified by 'test_batch')
test_batch <- "G6"
counts_batch <- assay(rds_obj,"counts")[, batch==test_batch]
cond_batch <- as.factor(as.character(group[batch==test_batch]))

y <- DGEList(counts=counts_batch, group=as.factor(cond_batch))
y <- calcNormFactors(y, method="TMM")
design <- model.matrix(~as.factor(cond_batch))
y <- estimateDisp(y, design)
fit <- glmQLFit(y, design)
qlf <- glmQLFTest(fit, coef=2)
de_res <- topTags(qlf, n=nrow(rds_obj))
 
up_genes <- rownames(de_res)[de_res$table$logFC >= 0]  # all up-regulated genes (descending order in terms of p value)
head_up_genes <- head(up_genes, n=50)  # top-50 significant up-regulated genes
tail_up_genes <- tail(up_genes, n=50)  # 50 least significant up-regulated genes

cond_up_inds <- up_genes

down_genes <- rownames(de_res)[de_res$table$logFC < 0]  # all down-regulated genes
head_down_genes <- head(down_genes, n=50)  # top-50 significant down-regulated genes
tail_down_genes <- tail(down_genes, n=50)  # 50 least significant down-regulated genes

cond_down_inds <- down_genes
```

```{r, echo=FALSE, message=FALSE}
# changes in terms of mean
cat("######  Changes in terms of average gene counts  ######\n")
mean_cts_mat <- matrix(NA, nrow=2, ncol=3, 
                       dimnames=list(c("Non-progressor", "Progressor"), c("top50 up", "top50 down", "100 least DE")))
mean_cts_mat["Non-progressor", "top50 up"] <- mean(counts_batch[head_up_genes, cond_batch=="Non-progressor"])
mean_cts_mat["Progressor", "top50 up"] <- mean(counts_batch[head_up_genes, cond_batch=="Progressor"])
mean_cts_mat["Non-progressor", "top50 down"] <- mean(counts_batch[head_down_genes, cond_batch=="Non-progressor"])
mean_cts_mat["Progressor", "top50 down"] <- mean(counts_batch[head_down_genes, cond_batch=="Progressor"])
mean_cts_mat["Non-progressor", "100 least DE"] <- mean(counts_batch[c(tail_up_genes, tail_down_genes), cond_batch=="Non-progressor"])
mean_cts_mat["Progressor", "100 least DE"] <- mean(counts_batch[c(tail_up_genes, tail_down_genes), cond_batch=="Progressor"])
print(round(mean_cts_mat,1))

# changes in terms of (log) fold change
cat("\n######  Range of fold change across biological conditions  ######\n")
FC_mat <- matrix(NA, nrow=4, ncol=3, 
                 dimnames=list(c("min FC", "median FC", "mean FC", "max FC"), 
                               c("top50 up", "top50 down", "100 least DE")))
FC_mat[, "top50 up"] <- sapply(c(min, median, mean, max), function(ff){ff(exp(de_res$table[head_up_genes, "logFC"]))})
FC_mat[, "top50 down"] <- sapply(c(min, median, mean, max), function(ff){ff(exp(de_res$table[head_down_genes, "logFC"]))})
FC_mat[, "100 least DE"] <- sapply(c(min, median, mean, max), function(ff){
  ff(exp(de_res$table[c(tail_up_genes, tail_down_genes), "logFC"]))
})
print(round(FC_mat,3))

cat("\n######  Range of Log(fold change) across biological conditions  ######\n")
logFC_mat <- log(FC_mat)
rownames(logFC_mat) <- c("min logFC", "median logFC", "mean logFC", "max logFC")
print(round(logFC_mat,3))


# visualize change in terms of (log) fold change
logFC_df <- data.frame(top50up=de_res$table[head_up_genes, "logFC"], top50down=de_res$table[head_down_genes, "logFC"])
FC_df <- exp(logFC_df)
p_logFC <- ggplot(melt(logFC_df), aes(x=variable, y=value)) +
  geom_boxplot() +
  labs(y="logFC") +
  theme(axis.title.x=element_blank())
p_FC <- ggplot(melt(FC_df), aes(x=variable, y=value)) +
  geom_boxplot() +
  labs(y="FC") +
  theme(axis.title.x=element_blank())
grid.arrange(p_logFC, p_FC, ncol=2)
```

```{r, echo=FALSE, results='hide'}
rm(cond_batch, counts_batch, up_genes, head_up_genes, tail_up_genes, 
   down_genes, head_down_genes, tail_down_genes, y, design, fit, qlf, de_res, 
   mean_cts_mat, FC_mat, p_FC, p_logFC, FC_df, logFC_df, logFC_mat)
```

**Conclusions:**

+ Biological fold change comparing average counts is about 1.2-1.8. 
+ Gene-wise fold change has a wide range (median 4, mean 21).


#### Change WRT batch

We are also interested to see how the average (mean) gene counts change across batches (i.e. **degree of mean batch effect**). To explore this mean batch effect, we perform DE with regard to the batch variable. To rule out the impact of biological condition, we perform this analysis within **Non-progressors** only, and compare **Africa & G6 studies**. Note that this DE is not true DE - it's testing if there is a significant mean difference between two batches. 

```{r, results="hide"}
####  Differential genes relative to batch 
####  (within condition 0 / condition specified by 'test_cond')
rds_subset <- rds_obj[, rds_obj$Dataset %in% c("Africa", "G6")]
test_cond <- "Non-progressor"
counts_cond <- assay(rds_subset, "counts")[, rds_subset$Label==test_cond]
batch_cond <- as.factor(as.character(rds_subset$SequencingBatch[rds_subset$Label==test_cond]))

y <- DGEList(counts=counts_cond, group=as.factor(batch_cond))
y <- calcNormFactors(y, method="TMM")
design <- model.matrix(~as.factor(batch_cond))
y <- estimateDisp(y, design)
fit <- glmQLFit(y, design)
qlf <- glmQLFTest(fit, coef=2)
de_res <- topTags(qlf, n=nrow(rds_subset))

up_genes <- rownames(de_res)[de_res$table$logFC >= 0]  # all up-regulated genes
head_up_genes <- head(up_genes, n=50)  # top-50 up-regulated genes
tail_up_genes <- tail(up_genes, n=50)  # 50 least up-regulated genes

batch_up_inds <- up_genes

down_genes <- rownames(de_res)[de_res$table$logFC < 0]  # all down-regulated genes
head_down_genes <- head(down_genes, n=50)  # top-50 down-regulated genes
tail_down_genes <- tail(down_genes, n=50)  # 50 least down-regulated genes

batch_down_inds <- down_genes
```

```{r, echo=FALSE}
# changes in terms of mean
cat("######  Batch differences in terms of mean gene counts  ######\n")
mean_cts_mat <- matrix(NA, nrow=2, ncol=3, 
                       dimnames=list(c("Africa", "G6"), c("top50 up", "top50 down", "100 least changed")))
mean_cts_mat["Africa", "top50 up"] <- mean(counts_cond[head_up_genes, batch_cond=="Africa"])
mean_cts_mat["G6", "top50 up"] <- mean(counts_cond[head_up_genes, batch_cond=="G6"])
mean_cts_mat["Africa", "top50 down"] <- mean(counts_cond[head_down_genes, batch_cond=="Africa"])
mean_cts_mat["G6", "top50 down"] <- mean(counts_cond[head_down_genes, batch_cond=="G6"])
mean_cts_mat["Africa", "100 least changed"] <- mean(counts_cond[c(tail_up_genes, tail_down_genes), batch_cond=="Africa"])
mean_cts_mat["G6", "100 least changed"] <- mean(counts_cond[c(tail_up_genes, tail_down_genes), batch_cond=="G6"])
print(round(mean_cts_mat,2))

# changes in terms of fold change
cat("\n######  Range of fold change between batches  ######\n")
cat("- All genes - FC:\n")
cat(round(range(exp(de_res$table$logFC)), 3)) 

cat("\n\n- All genes - log(FC):\n")
cat(round(range(de_res$table$logFC), 3)) 

cat("\n\n- Top genes - FC:\n")
FC_mat <- matrix(NA, nrow=4, ncol=3, 
                 dimnames=list(c("min FC", "median FC", "mean FC", "max FC"), 
                               c("top50 up", "top50 down", "100 least DE")))
FC_mat[, "top50 up"] <- sapply(c(min, median, mean, max), function(ff){ff(exp(de_res$table[head_up_genes, "logFC"]))})
FC_mat[, "top50 down"] <- sapply(c(min, median, mean, max), function(ff){ff(exp(de_res$table[head_down_genes, "logFC"]))})
FC_mat[, "100 least DE"] <- sapply(c(min, median, mean, max), function(ff){
  ff(exp(de_res$table[c(tail_up_genes, tail_down_genes), "logFC"]))
})
print(round(FC_mat,3))

cat("\n- Top genes - log(FC):\n")
logFC_mat <- log(FC_mat)
rownames(logFC_mat) <- c("min logFC", "median logFC", "mean logFC", "max logFC")
print(round(logFC_mat,3))

```

```{r, echo=FALSE, results='hide'}
rm(batch_cond, counts_cond, up_genes, head_up_genes, tail_up_genes, 
   down_genes, head_down_genes, tail_down_genes, y, design, fit, qlf, de_res, 
   mean_cts_mat, FC_mat, logFC_mat)
```

**Conclusions:**

+ Mean batch effect fold change comparing average counts ranges from 1.18-3.56. 
+ Gene-wise mean batch effect fold change has a wide range (mean 4.4, median 52.6).


### Dispersion

Aside from average gene counts, we are interested in comparing dispersions in the batches. **Difference in the dispersions across batches suggests variance batch effect**. We visualize the gene-wise dispersion estimates from edgeR:

```{r, echo=FALSE}
####  Dispersion differences across batches 
y_batch1 <- DGEList(counts=assay(rds_obj,"counts")[, batch=="Africa"], group=as.factor(as.character(group[batch=="Africa"])))
y_batch1 <- calcNormFactors(y_batch1, method="TMM")
design_batch1 <- model.matrix(~as.factor(as.character(group[batch=="Africa"])))
y_batch1 <- estimateDisp(y_batch1, design_batch1)

y_batch2 <- DGEList(counts=assay(rds_obj,"counts")[, batch=="G6"], group=as.factor(as.character(group[batch=="G6"])))
y_batch2 <- calcNormFactors(y_batch2, method="TMM")
design_batch2 <- model.matrix(~as.factor(as.character(group[batch=="G6"])))
y_batch2 <- estimateDisp(y_batch2, design_batch2)

y_batch3 <- DGEList(counts=assay(rds_obj,"counts")[, batch=="India"], group=as.factor(as.character(group[batch=="India"])))
y_batch3 <- calcNormFactors(y_batch3, method="TMM")
design_batch3 <- model.matrix(~as.factor(as.character(group[batch=="India"])))
y_batch3 <- estimateDisp(y_batch3, design_batch3)

y_batch4 <- DGEList(counts=assay(rds_obj,"counts")[, batch=="Brazil_1"], group=as.factor(as.character(group[batch=="Brazil_1"])))
y_batch4 <- calcNormFactors(y_batch4, method="TMM")
design_batch4 <- model.matrix(~as.factor(as.character(group[batch=="Brazil_1"])))
y_batch4 <- estimateDisp(y_batch4, design_batch4)

y_batch5 <- DGEList(counts=assay(rds_obj,"counts")[, batch=="Brazil_2"], group=as.factor(as.character(group[batch=="Brazil_2"])))
y_batch5 <- calcNormFactors(y_batch5, method="TMM")
#design_batch5 <- model.matrix(~as.factor(as.character(group[batch=="Brazil_2"])))
y_batch5 <- estimateDisp(y_batch5)

disp_df <- data.frame(Africa=y_batch1$tagwise.dispersion, G6=y_batch2$tagwise.dispersion, 
                      India=y_batch3$tagwise.dispersion, Brazil_1=y_batch4$tagwise.dispersion,
                      Brazil_2=y_batch5$tagwise.dispersion)

cat("####  Statistics about dispersion estimates from the two batches  ####\n")
disp_stats <- lapply(c(min, median, mean, max), function(ff){apply(disp_df, 2, ff)})
disp_stats <- do.call(rbind, disp_stats)
rownames(disp_stats) <- c("min disp.", "median disp.", "mean disp.", "max disp.")
print(round(disp_stats, 4))
```

```{r, echo=FALSE, message=FALSE}
disp_df_mlt <- melt(disp_df)
ggplot(disp_df_mlt, aes(x=variable, y=value)) +
  geom_violin() +
  labs(x="Batch", y="Estimated gene-wise dispersion", 
       title="Comparison of estimated dispersion across batches") +
  stat_summary(fun.y=mean, colour="darkred", geom="point", shape=18, size=3) #+
  #scale_y_continuous(limits=c(0, 1))
```

**Conclusions:**

+ Mean dispersion in the Africa / Batch 1 in the Brazil study is about 10 times of those in the India study. 
+ We see huge max gene-wise dispersions, possibly caused by 0 counts.


### Variance

**Median** gene-wise variance: 

```{r, echo=FALSE}
var_tb <- matrix(NA, nrow=2, ncol=3, 
                 dimnames=list(c("Non-progressor", "Progressor"), c("Africa", "G6", "Brazil1")))
var_tb["Non-progressor", "Africa"] <- median(rowVars(cts[, batch=="Africa"&group=="Non-progressor"]))
var_tb["Progressor", "Africa"] <- median(rowVars(cts[, batch=="Africa"&group=="Progressor"]))
var_tb["Non-progressor", "G6"] <- median(rowVars(cts[, batch=="G6"&group=="Non-progressor"]))
var_tb["Progressor", "G6"] <- median(rowVars(cts[, batch=="G6"&group=="Progressor"]))
var_tb["Non-progressor", "Brazil1"] <- median(rowVars(cts[, batch=="Brazil_1"&group=="Non-progressor"]))
var_tb["Progressor", "Brazil1"] <- median(rowVars(cts[, batch=="Brazil_1"&group=="Progressor"]))
print(round(var_tb, 2))
```

Scaled by that of Africa:

```{r, echo=FALSE}
var_tb_scaled <- apply(var_tb, 2, function(x){x/var_tb[,"Africa"]})
print(round(var_tb_scaled, 2))
```


## Impact of 0 counts on gene distributions

We have seen above that over half of genes in the dataset contain some portions of zeros. Having 0 counts can have a good impact on mean and dispersion estimation. We evaluate such impact in this section, and see how NB distribution fits to non-zero portion of genes. We first take the non-zero portion of the genes.

### Fold change in mean in non-zero portion {.tabset}

```{r, echo=FALSE}
# take the non-zero portion in each gene
batch_sep <- group_sep <- nonzero_cts <- studies_sep <- list()
for(i in 1:nrow(cts)){
  nonzero_ind <- which(cts[i, ]!=0)
  batch_sep[[i]] <- as.character(batch[nonzero_ind])
  group_sep[[i]] <- as.character(group[nonzero_ind])
  nonzero_cts[[i]] <- cts[i, nonzero_ind]
  studies_sep[[i]] <- as.character(rds_obj$Dataset[nonzero_ind])
}
# sanity check
if(!identical(sapply(nonzero_cts, length), as.integer(rowSums(cts!=0)))){stop("Error in taking non-zero portions!")}

names(nonzero_cts) <- rownames(cts); names(batch_sep) <- rownames(cts); names(group_sep) <- rownames(cts)
```

####  Change WRT condition in G6

Within **G6**, we compare mean (here refers to non-zero mean) across biological condition groups. Specifically, **we saved the previous DE analysis results using edgeR, on counts containing zeros**. We call all genes with $\log(FC) >= 0$ as "up" genes, and $\log(FC) < 0$ as "down" genes. The top 50 "up" and "down" genes are still determined by the previous DE containing zeros. 

```{r, echo=FALSE}
# take G6
group_sep_batch1 <- nonzero_cts_batch1 <- list()
for(i in 1:length(nonzero_cts)){
  group_sep_batch1[[i]] <- group_sep[[i]][batch_sep[[i]]=="G6"]
  nonzero_cts_batch1[[i]] <- nonzero_cts[[i]][batch_sep[[i]]=="G6"]
}
names(nonzero_cts_batch1) <- names(group_sep_batch1) <- rownames(cts)

# split into controls & cases
nonzero_cts_b1_ctrls <- nonzero_cts_b1_cases <- list()
for(j in 1:length(nonzero_cts_batch1)){
  nonzero_cts_b1_ctrls[[j]] <- nonzero_cts_batch1[[j]][group_sep_batch1[[j]]=="Non-progressor"]
  nonzero_cts_b1_cases[[j]] <- nonzero_cts_batch1[[j]][group_sep_batch1[[j]]=="Progressor"]
}#identical(sapply(nonzero_cts_b1_ctrls, length) + sapply(nonzero_cts_b1_cases, length), sapply(nonzero_cts_batch1,length))
names(nonzero_cts_b1_ctrls) <- names(nonzero_cts_b1_cases) <- rownames(cts)

# top 50 up and down indices
top_up_ind <- head(cond_up_inds, 50)
top_down_ind <- head(cond_down_inds, 50)
least_de_ind <- c(tail(cond_up_inds, 50), tail(cond_down_inds, 50))

# calculate mean in top up, down and least DE
cat("######  Changes in MEAN non-zero gene counts  ######\n")
mean_nonzerocts_mat <- matrix(NA, nrow=2, ncol=3, 
                              dimnames=list(c("Non-progressor", "Progressor"), c("top50 up", "top50 down", "100 least DE")))
mean_nonzerocts_mat["Non-progressor", "top50 up"] <- mean(do.call(c, nonzero_cts_b1_ctrls[top_up_ind]))
mean_nonzerocts_mat["Progressor", "top50 up"] <- mean(do.call(c, nonzero_cts_b1_cases[top_up_ind]))
mean_nonzerocts_mat["Non-progressor", "top50 down"] <- mean(do.call(c, nonzero_cts_b1_ctrls[top_down_ind]))
mean_nonzerocts_mat["Progressor", "top50 down"] <- mean(do.call(c, nonzero_cts_b1_cases[top_down_ind]))
mean_nonzerocts_mat["Non-progressor", "100 least DE"] <- mean(do.call(c, nonzero_cts_b1_ctrls[least_de_ind]))
mean_nonzerocts_mat["Progressor", "100 least DE"] <- mean(do.call(c, nonzero_cts_b1_cases[least_de_ind]))
print(round(mean_nonzerocts_mat,1))
```

```{r,echo=FALSE}
cat("######  Changes in MEDIAN non-zero gene counts  ######\n")
median_nonzerocts_mat <- matrix(NA, nrow=2, ncol=3, 
                              dimnames=list(c("Non-progressor", "Progressor"), c("top50 up", "top50 down", "100 least DE")))
median_nonzerocts_mat["Non-progressor", "top50 up"] <- median(do.call(c, nonzero_cts_b1_ctrls[top_up_ind]))
median_nonzerocts_mat["Progressor", "top50 up"] <- median(do.call(c, nonzero_cts_b1_cases[top_up_ind]))
median_nonzerocts_mat["Non-progressor", "top50 down"] <- median(do.call(c, nonzero_cts_b1_ctrls[top_down_ind]))
median_nonzerocts_mat["Progressor", "top50 down"] <- median(do.call(c, nonzero_cts_b1_cases[top_down_ind]))
median_nonzerocts_mat["Non-progressor", "100 least DE"] <- median(do.call(c, nonzero_cts_b1_ctrls[least_de_ind]))
median_nonzerocts_mat["Progressor", "100 least DE"] <- median(do.call(c, nonzero_cts_b1_cases[least_de_ind]))
print(round(median_nonzerocts_mat,1))
```

**Conclusions: ** 

+ Biological fold change (in non-zero portion) comparing average counts is about 1.2-1.8 (mean).
+ O counts did not strongly affect biological signal in terms of average counts.


####  Change WRT batch in non-progressors

Similarly, we compare non-zero mean across batches within "Non-progressors" of Africa and G6 studies.

```{r, echo=FALSE}
# take controls
batch_sep_cond0 <- nonzero_cts_cond0 <- list()
for(i in 1:length(nonzero_cts)){
  batch_sep_cond0[[i]] <- batch_sep[[i]][group_sep[[i]]=="Non-progressor"]
  nonzero_cts_cond0[[i]] <- nonzero_cts[[i]][group_sep[[i]]=="Non-progressor"]
}
names(batch_sep_cond0) <- names(nonzero_cts_cond0) <- rownames(cts)

# split into batch 1 & 2
nonzero_cts_cond0_batch1 <- nonzero_cts_cond0_batch2 <- list()
for(j in 1:length(nonzero_cts)){
  nonzero_cts_cond0_batch1[[j]] <- nonzero_cts_cond0[[j]][batch_sep_cond0[[j]]=="Africa"]
  nonzero_cts_cond0_batch2[[j]] <- nonzero_cts_cond0[[j]][batch_sep_cond0[[j]]=="G6"]
}
names(nonzero_cts_cond0_batch1) <- names(nonzero_cts_cond0_batch2) <- rownames(cts)

# top 50 up and down indices
top_up_ind <- head(batch_up_inds, 50)
top_down_ind <- head(batch_down_inds, 50)
least_de_ind <- c(tail(batch_up_inds, 50), tail(batch_down_inds, 50))

# calculate mean in top up, down and least DE
cat("######  Changes in MEAN non-zero gene counts  ######\n")
mean_nonzerocts_mat <- matrix(NA, nrow=2, ncol=3, 
                              dimnames=list(c("Africa", "G6"), c("top50 up", "top50 down", "100 least DE")))
mean_nonzerocts_mat["Africa", "top50 up"] <- mean(do.call(c, nonzero_cts_cond0_batch1[top_up_ind]))
mean_nonzerocts_mat["G6", "top50 up"] <- mean(do.call(c, nonzero_cts_cond0_batch2[top_up_ind]))
mean_nonzerocts_mat["Africa", "top50 down"] <- mean(do.call(c, nonzero_cts_cond0_batch1[top_down_ind]))
mean_nonzerocts_mat["G6", "top50 down"] <- mean(do.call(c, nonzero_cts_cond0_batch2[top_down_ind]))
mean_nonzerocts_mat["Africa", "100 least DE"] <- mean(do.call(c, nonzero_cts_cond0_batch1[least_de_ind]))
mean_nonzerocts_mat["G6", "100 least DE"] <- mean(do.call(c, nonzero_cts_cond0_batch2[least_de_ind]))
print(round(mean_nonzerocts_mat,1))
```

```{r, echo=FALSE}
# calculate median in top up, down and least DE
cat("######  Changes in MEDIAN non-zero gene counts  ######\n")
median_nonzerocts_mat <- matrix(NA, nrow=2, ncol=3, 
                                dimnames=list(c("Africa", "G6"), c("top50 up", "top50 down", "100 least DE")))
median_nonzerocts_mat["Africa", "top50 up"] <- median(do.call(c, nonzero_cts_cond0_batch1[top_up_ind]))
median_nonzerocts_mat["G6", "top50 up"] <- median(do.call(c, nonzero_cts_cond0_batch2[top_up_ind]))
median_nonzerocts_mat["Africa", "top50 down"] <- median(do.call(c, nonzero_cts_cond0_batch1[top_down_ind]))
median_nonzerocts_mat["G6", "top50 down"] <- median(do.call(c, nonzero_cts_cond0_batch2[top_down_ind]))
median_nonzerocts_mat["Africa", "100 least DE"] <- median(do.call(c, nonzero_cts_cond0_batch1[least_de_ind]))
median_nonzerocts_mat["G6", "100 least DE"] <- median(do.call(c, nonzero_cts_cond0_batch2[least_de_ind]))
print(round(median_nonzerocts_mat,1))
```

**Conclusions:** 

+ Mean batch effect fold change (non-zero portion) comparing average counts ranges from 1.16-3.3 (mean). 
+ O counts did not strongly affect mean batch effect in terms of average counts.


### Dispersion with 0 counts removed 

Dispersion estimates are more likely impacted by 0 counts. We re-estimate the dispersions in each batch, using non-zero portion of each gene. Then we compare the dispersion estimates across batches.

```{r, echo=FALSE}
load(file.path(output_dir, "disps_nonzero.RData"))
disp_nonzero_lst <- list(Africa=disp_nonzero_africa, G6=disp_nonzero_G6,
                         India=disp_nonzero_india, Brazil_1=disp_nonzero_bzl1, 
                         Brazil_2=disp_nonzero_bzl2)
cat("####  Statistics about dispersion estimates (from NON-ZERO portions) from the two batches  ####\n")
disp_stats <- lapply(c(min, median, mean, max), function(ff){sapply(disp_nonzero_lst, ff)})
disp_stats <- do.call(rbind, disp_stats)
rownames(disp_stats) <- c("min disp.", "median disp.", "mean disp.", "max disp.")
print(round(disp_stats, 4))
```

```{r, echo=FALSE, message=FALSE}
disp_nonzero_df <- data.frame(Batch=c(rep("Africa", length(disp_nonzero_africa)), 
                                      rep("G6", length(disp_nonzero_G6)),
                                      rep("India", length(disp_nonzero_india)),
                                      rep("Brazil_1", length(disp_nonzero_bzl1)),
                                      rep("Brazil_2", length(disp_nonzero_bzl2))),
                              value=do.call(c, disp_nonzero_lst))
ggplot(disp_nonzero_df , aes(x=Batch, y=value)) +
  geom_violin() +
  labs(x="Batch", y="Estimated gene-wise dispersion on NON-ZERO counts", 
       title="Comparison of estimated dispersion across batches") +
  stat_summary(fun.y=mean, colour="darkred", geom="point", shape=18, size=3) #+
  #scale_y_continuous(limits=c(0, 1))
```

**Conclusions:**

+ Mean dispersion of Brazil batch 1 is roughly 3.4 times that of India.
+ The absolute value of max gene-wise dispersion are reduced after 0 counts are removed.
+ In this report, dispersion are calculated by different methods before and after removing 0 counts. **Not directly comparable.**
+ We show in another analysis (*TB_run_nonzero_disps.R*) that **the difference in dispersion is reduced after removing 0s**.


### Variance with 0 counts removed 

**Median** gene-wise variance after removing zeros: 

```{r, echo=FALSE}
var_tb_rmzero <- matrix(NA, nrow=2, ncol=3, 
                        dimnames=list(c("Non-progressor", "Progressor"), c("Africa", "G6", "Brazil1")))
cts_rmzero <- cts; cts_rmzero[cts_rmzero==0] <- NA
var_tb_rmzero["Non-progressor", "Africa"] <- median(rowVars(cts_rmzero[, batch=="Africa"&group=="Non-progressor"], na.rm=TRUE), na.rm=TRUE)
var_tb_rmzero["Progressor", "Africa"] <- median(rowVars(cts_rmzero[, batch=="Africa"&group=="Progressor"], na.rm=TRUE), na.rm=TRUE)
var_tb_rmzero["Non-progressor", "G6"] <- median(rowVars(cts_rmzero[, batch=="G6"&group=="Non-progressor"], na.rm=TRUE), na.rm=TRUE)
var_tb_rmzero["Progressor", "G6"] <- median(rowVars(cts_rmzero[, batch=="G6"&group=="Progressor"], na.rm=TRUE), na.rm=TRUE)
var_tb_rmzero["Non-progressor", "Brazil1"] <- median(rowVars(cts_rmzero[, batch=="Brazil_1"&group=="Non-progressor"], na.rm=TRUE), na.rm=TRUE)
var_tb_rmzero["Progressor", "Brazil1"] <- median(rowVars(cts_rmzero[, batch=="Brazil_1"&group=="Progressor"], na.rm=TRUE), na.rm=TRUE)
print(round(var_tb_rmzero, 2))
```

Scaled by that of Africa:

```{r, echo=FALSE}
var_tb_rmzero_scaled <- apply(var_tb_rmzero, 2, function(x){x/var_tb_rmzero[,"Africa"]})
print(round(var_tb_rmzero_scaled, 2))
```


### Is there a difference in zero portions across batches?

Another question related to batch effect, is whether there is any difference in the percentage of 0 counts in a gene between two batches. 

```{r, echo=FALSE, message=FALSE}
n_zeros_batch_africa <- apply(cts[, batch=="Africa"], 1, function(x){length(which(x==0))})
percent_zeros_batch_africa <- n_zeros_batch_africa / sum(batch=="Africa")
n_zeros_batch_G6 <- apply(cts[, batch=="G6"], 1, function(x){length(which(x==0))})
percent_zeros_batch_G6 <- n_zeros_batch_G6 / sum(batch=="G6")
n_zeros_batch_india <- apply(cts[, batch=="India"], 1, function(x){length(which(x==0))})
percent_zeros_batch_india <- n_zeros_batch_india / sum(batch=="India")
n_zeros_batch_bzl1 <- apply(cts[, batch=="Brazil_1"], 1, function(x){length(which(x==0))})
percent_zeros_batch_bzl1 <- n_zeros_batch_bzl1 / sum(batch=="Brazil_1")
n_zeros_batch_bzl2 <- apply(cts[, batch=="Brazil_2"], 1, function(x){length(which(x==0))})
percent_zeros_batch_bzl2 <- n_zeros_batch_bzl2 / sum(batch=="Brazil_2")

percent_zeros_batch_df <- data.frame(Africa=percent_zeros_batch_africa, G6=percent_zeros_batch_G6,
                                     India=percent_zeros_batch_india, Brazil_1=percent_zeros_batch_bzl1,
                                     Brazil_2=percent_zeros_batch_bzl2)
percent_zeros_batch_df_mlt <- melt(percent_zeros_batch_df, variable.name="Batch")
ggplot(percent_zeros_batch_df_mlt, aes(x=Batch, y=value)) +
  geom_violin() +
  labs(y="Percentage of 0 counts in a single gene") +
  theme(axis.title.x=element_blank())
```

**Conclusions:**

+ There isn't an obvious difference in zero fractions between batches (aside from Brazil batch 2 which is too small).


### Goodness of fit for NB distribution to non-zero portion of genes

It is difficult to evaluate quantitatively the goodness-of-fit of NB distribution on data. Instead, we generate Q-Q plots for 100 randomly selected genes, comparing data with NB distribution using ML estimated parameters. We generate these plots **within G6/Africa, for cases and control samples separately**. Figures are stored under {output_dir}/qqplots_{Study}/.

```{r, echo=FALSE, results="hide"}
if(!dir.exists(file.path(output_dir, "qqplots_G6"))){dir.create(file.path(output_dir, "qqplots_G6"))}
set.seed(1234)
N_sel_nb <- 100
chosen_id_nb <- sample(1:nrow(rds_obj), N_sel_nb, replace=FALSE)
cts_subset_nb <- cts[chosen_id_nb, ]
for(i in 1:N_sel_nb){
  # case samples
  curr_cts <- cts_subset_nb[i, batch=="G6" & group=="Progressor"]
  curr_cts <- curr_cts[curr_cts!=0]
  if(length(unique(curr_cts)) <= 3 | var(curr_cts)==0){next}
  df <- data.frame(y=curr_cts)
  params <- fitdistr(df$y, "negative binomial")$estimate
  
  png(paste0(output_dir, "/qqplots_G6/gene", i, "_cases.png"), 
      width=5, height=5, units="in", res=300)
  print(ggplot(df, aes(sample=y)) +
          stat_qq(distribution=qnbinom, dparams=params) + 
          stat_qq_line(distribution=qnbinom, dparams=params) +
          labs(title="Q-Q plot: G6 Progressors"))
  dev.off()
  rm(curr_cts, df, params)
  
  # control samples
  curr_cts <- cts_subset_nb[i, batch=="G6" & group=="Non-progressor"]
  curr_cts <- curr_cts[curr_cts!=0]
  if(length(unique(curr_cts)) <= 3 | var(curr_cts)==0){next}
  df <- data.frame(y=curr_cts)
  params <- fitdistr(df$y, "negative binomial")$estimate
  
  png(paste0(output_dir, "/qqplots_G6/gene", i, "_controls.png"), 
      width=5, height=5, units="in", res=300)
  print(ggplot(df, aes(sample=y)) +
          stat_qq(distribution=qnbinom, dparams=params) + 
          stat_qq_line(distribution=qnbinom, dparams=params) +
          labs(title="Q-Q plot: G6 Non-progressors"))
  dev.off()
}
```

```{r, echo=FALSE, results="hide"}
if(!dir.exists(file.path(output_dir, "qqplots_africa"))){dir.create(file.path(output_dir, "qqplots_africa"))}
set.seed(1234)
N_sel_nb <- 100
chosen_id_nb <- sample(1:nrow(rds_obj), N_sel_nb, replace=FALSE)
cts_subset_nb <- cts[chosen_id_nb, ]
for(i in 1:N_sel_nb){
  # case samples
  curr_cts <- cts_subset_nb[i, batch=="Africa" & group=="Progressor"]
  curr_cts <- curr_cts[curr_cts!=0]
  if(length(unique(curr_cts)) <= 3 | var(curr_cts)==0){next}
  df <- data.frame(y=curr_cts)
  params <- fitdistr(df$y, "negative binomial")$estimate
  
  png(paste0(output_dir, "/qqplots_africa/gene", i, "_cases.png"), 
      width=5, height=5, units="in", res=300)
  print(ggplot(df, aes(sample=y)) +
          stat_qq(distribution=qnbinom, dparams=params) + 
          stat_qq_line(distribution=qnbinom, dparams=params) +
          labs(title="Q-Q plot: Africa Progressors"))
  dev.off()
  rm(curr_cts, df, params)
  
  # control samples
  curr_cts <- cts_subset_nb[i, batch=="Africa" & group=="Non-progressor"]
  curr_cts <- curr_cts[curr_cts!=0]
  if(length(unique(curr_cts)) <= 3 | var(curr_cts)==0){next}
  df <- data.frame(y=curr_cts)
  params <- fitdistr(df$y, "negative binomial")$estimate
  
  png(paste0(output_dir, "/qqplots_africa/gene", i, "_controls.png"), 
      width=5, height=5, units="in", res=300)
  print(ggplot(df, aes(sample=y)) +
          stat_qq(distribution=qnbinom, dparams=params) + 
          stat_qq_line(distribution=qnbinom, dparams=params) +
          labs(title="Q-Q plot: Africa Non-progressors"))
  dev.off()
}
```

## Session info 

```{r, echo=FALSE}
sessionInfo()
```



## References
