---
title: "Application of ComBat-seq on simulated datasets"
subtitle: "Levels of dispersion differences and confounding"
author: Yuqing Zhang
date: "`r Sys.Date()`"
output:
  html_notebook:
    theme: united
    toc: yes
---

## About this simulation

From previous analysis (*"powerdown_unbalanced"*), it appears that ComBat-seq works when there is variance batch effect **AND** when there is some level of confounding between batch and condition (i.e. in unbalanced study design). We dig deeper in these two factors.


## Simulation setup

* Mean fold changes: Biological - 2, Batch - 1.5 
* Dispersion: Batch 1 - 0.15, Batch 2 - 0.15***dispersion fold level**
* Study design: **Different levels of confounding**
* Reads per gene: Median 120 (coverage 5)
* Total number of samples: 20 (*we look at N=60 later in comparison to see if patterns hold*)

Note that the fixed parameters are consistent with (i.e. within the range of) my observations from real datasets (See explore_real_data for details). 

Two factors under examination are dispersion fold change and level of confounding. Confounding level is measured by **how many percentage of samples in batch 1 is control samples**. 2 batches have the same size, but in batch 2, the number of controls and cases are reversed compared to batch 1. Grids for the two factors:

* Dispersion level: 1 (no dispersion difference), 2, 3, 4, 5
* Confounding: 0.5 (no confounding), 0.4, 0.3, 0.2

```{r, echo=FALSE, results='hide'}
rm(list=ls())
sapply(c("ggplot2", "gridExtra", "reshape2", "ggpubr"), require, character.only=TRUE)
results_dir <- "~/Google Drive/ComBat_seq/DE_disp_confounding_base"
#results_dir <- "~/yuqingz/ComBat_seq/DE_disp_confounding_base"
N_samples <- 60

disp_level_vec <- 1:5
confounding_level_vec <- seq(from=0.5,to=0.2,by=-0.1)
sel_method_list <- list(edgeR=c("BaseIndi.edgeR", #"BaseQuant.edgeR", 
                                "Batch.edgeR", "OneStep.edgeR", "ComBat.lm", 
                                "RUVseq.edgeR", "SVAseq.edgeR", "ComBatseq.edgeR"),
                        DESeq2=c("BaseIndi.DESeq2", #"BaseQuant.DESeq2", 
                                 "Batch.DESeq2", "OneStep.DESeq2", "ComBat.lm", 
                                 "RUVseq.DESeq2", "SVAseq.DESeq2", "ComBatseq.DESeq2"))
```


## Results 

### TPR vs FPR {.tabset}

```{r, echo=FALSE, results='hide'}
fpr_all_lst <- tpr_all_lst <- list()
for(i in 1:length(confounding_level_vec)){
  fpr_all_lst[[i]] <- tpr_all_lst[[i]] <- list()
  for(j in 1:length(disp_level_vec)){
    exp_name <- paste0("simDCbase_N", N_samples, "_dispFC", disp_level_vec[j],
                       "_percent", gsub(".","", confounding_level_vec[i], fixed=T))
    tpr_res <- read.csv(file.path(results_dir, sprintf('tpr_%s.csv', exp_name)))
    fpr_res <- read.csv(file.path(results_dir, sprintf('fpr_%s.csv', exp_name)))
    
    tpr_res$BaseQuant.edgeR <- NULL; tpr_res$BaseQuant.DESeq2 <- NULL
    fpr_res$BaseQuant.edgeR <- NULL; fpr_res$BaseQuant.DESeq2 <- NULL
    
    tpr_all_lst[[i]][[j]] <- as.list(colMeans(tpr_res, na.rm=TRUE))
    fpr_all_lst[[i]][[j]]  <- as.list(colMeans(fpr_res, na.rm=TRUE))
  }
  names(tpr_all_lst[[i]]) <- names(fpr_all_lst[[i]]) <- paste0("Disp", disp_level_vec)
}
names(tpr_all_lst) <- names(fpr_all_lst) <- paste0("Cnfnd", confounding_level_vec)

## All methods
tpr_merged <- melt(tpr_all_lst); colnames(tpr_merged) <- c("TPR", "Method", "Disp", "Cnfnd")
fpr_merged <- melt(fpr_all_lst); colnames(fpr_merged) <- c("FPR", "Method", "Disp", "Cnfnd")
tpr_merged$Method <- factor(tpr_merged$Method, levels=colnames(tpr_res))
fpr_merged$Method <- factor(fpr_merged$Method, levels=colnames(fpr_res))
tpr_merged$Cnfnd <- factor(tpr_merged$Cnfnd, levels=paste0("Cnfnd", confounding_level_vec))
fpr_merged$Cnfnd <- factor(fpr_merged$Cnfnd, levels=paste0("Cnfnd", confounding_level_vec))


## edgeR only
tpr_edgeR_lst <- lapply(tpr_all_lst, function(t1_lst){lapply(t1_lst, function(t2_lst){t2_lst[sel_method_list$edgeR]})})
fpr_edgeR_lst <- lapply(fpr_all_lst, function(f1_lst){lapply(f1_lst, function(f2_lst){f2_lst[sel_method_list$edgeR]})})

tpr_edgeR_merged <- melt(tpr_edgeR_lst); colnames(tpr_edgeR_merged) <- c("TPR", "Method", "Disp", "Cnfnd")
fpr_edgeR_merged <- melt(fpr_edgeR_lst); colnames(fpr_edgeR_merged) <- c("FPR", "Method", "Disp", "Cnfnd")
tpr_edgeR_merged$Method <- factor(tpr_edgeR_merged$Method, levels=sel_method_list$edgeR)
fpr_edgeR_merged$Method <- factor(fpr_edgeR_merged$Method, levels=sel_method_list$edgeR)
tpr_edgeR_merged$Cnfnd <- factor(tpr_edgeR_merged$Cnfnd, levels=paste0("Cnfnd", confounding_level_vec))
fpr_edgeR_merged$Cnfnd <- factor(fpr_edgeR_merged$Cnfnd, levels=paste0("Cnfnd", confounding_level_vec))


## DESeq2 only
tpr_DESeq2_lst <- lapply(tpr_all_lst, function(t1_lst){lapply(t1_lst, function(t2_lst){t2_lst[sel_method_list$DESeq2]})})
fpr_DESeq2_lst <- lapply(fpr_all_lst, function(f1_lst){lapply(f1_lst, function(f2_lst){f2_lst[sel_method_list$DESeq2]})})

tpr_DESeq2_merged <- melt(tpr_DESeq2_lst); colnames(tpr_DESeq2_merged) <- c("TPR", "Method", "Disp", "Cnfnd")
fpr_DESeq2_merged <- melt(fpr_DESeq2_lst); colnames(fpr_DESeq2_merged) <- c("FPR", "Method", "Disp", "Cnfnd")
tpr_DESeq2_merged$Method <- factor(tpr_DESeq2_merged$Method, levels=sel_method_list$DESeq2)
fpr_DESeq2_merged$Method <- factor(fpr_DESeq2_merged$Method, levels=sel_method_list$DESeq2)
tpr_DESeq2_merged$Cnfnd <- factor(tpr_DESeq2_merged$Cnfnd, levels=paste0("Cnfnd", confounding_level_vec))
fpr_DESeq2_merged$Cnfnd <- factor(fpr_DESeq2_merged$Cnfnd, levels=paste0("Cnfnd", confounding_level_vec))
```


#### Line plots {.tabset}

##### Facet by confounding levels

```{r, echo=FALSE}
## All methods
p_fpr <- ggplot(fpr_merged, aes(x=Disp, y=FPR, group=Method, color=Method)) +
  facet_grid(~Cnfnd) +
  geom_line() +
  geom_hline(yintercept=0.05, col="red", linetype="dashed") +
  labs(title="All") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
  #theme(axis.text.x=element_blank(), axis.title.x=element_blank()) 
p_tpr <- ggplot(tpr_merged, aes(x=Disp, y=TPR, group=Method, color=Method)) +
  facet_grid(~Cnfnd) +
  geom_line() +
  #labs(title="All") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_all <- ggarrange(plotlist=list(p_fpr, p_tpr), nrow=2, 
                   common.legend=TRUE, legend="bottom")#, heights=c(5.5,6))

## edgeR only
p_edgeR_fpr <- ggplot(fpr_edgeR_merged, aes(x=Disp, y=FPR, group=Method, color=Method)) +
  facet_grid(~Cnfnd) +
  geom_line() +
  geom_hline(yintercept=0.05, col="red", linetype="dashed") +
  labs(title="edgeR") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
  #theme(axis.text.x=element_blank(), axis.title.x=element_blank()) 
p_edgeR_tpr <- ggplot(tpr_edgeR_merged, aes(x=Disp, y=TPR, group=Method, color=Method)) +
  facet_grid(~Cnfnd) +
  geom_line() +
  #labs(title="edgeR") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_edgeR <- ggarrange(plotlist=list(p_edgeR_fpr, p_edgeR_tpr), nrow=2, 
                     common.legend=TRUE, legend="bottom")#, heights=c(5.5,6))

## DESeq2 only
p_DESeq2_fpr <- ggplot(fpr_DESeq2_merged, aes(x=Disp, y=FPR, group=Method, color=Method)) +
  facet_grid(~Cnfnd) +
  geom_line() +
  geom_hline(yintercept=0.05, col="red", linetype="dashed") +
  labs(title="DESeq2") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
  #theme(axis.text.x=element_blank(), axis.title.x=element_blank()) 
p_DESeq2_tpr <- ggplot(tpr_DESeq2_merged, aes(x=Disp, y=TPR, group=Method, color=Method)) +
  facet_grid(~Cnfnd) +
  geom_line() +
  #labs(title="DESeq2") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_DESeq2 <- ggarrange(plotlist=list(p_DESeq2_fpr, p_DESeq2_tpr), nrow=2, 
                      common.legend=TRUE, legend="bottom")#, heights=c(5.5,6))

#p <- ggarrange(plotlist=list(p_all, p_edgeR, p_DESeq2), nrow=3)
p <- ggarrange(plotlist=list(p_edgeR, p_DESeq2), nrow=2)
```

```{r, echo=FALSE, fig.width=5, fig.height=7}
p
```


##### Facet by dispersion levels

```{r, echo=FALSE}
## All methods
p_fpr <- ggplot(fpr_merged, aes(x=Cnfnd, y=FPR, group=Method, color=Method)) +
  facet_grid(~Disp) +
  geom_line() +
  geom_hline(yintercept=0.05, col="red", linetype="dashed") +
  labs(title="All") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
  #theme(axis.text.x=element_blank(), axis.title.x=element_blank()) 
p_tpr <- ggplot(tpr_merged, aes(x=Cnfnd, y=TPR, group=Method, color=Method)) +
  facet_grid(~Disp) +
  geom_line() +
  #labs(title="All") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_all <- ggarrange(plotlist=list(p_fpr, p_tpr), nrow=2, 
                   common.legend=TRUE, legend="bottom")#, heights=c(5.5,6))

## edgeR only
p_edgeR_fpr <- ggplot(fpr_edgeR_merged, aes(x=Cnfnd, y=FPR, group=Method, color=Method)) +
  facet_grid(~Disp) +
  geom_line() +
  geom_hline(yintercept=0.05, col="red", linetype="dashed") +
  labs(title="edgeR") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
  #theme(axis.text.x=element_blank(), axis.title.x=element_blank()) 
p_edgeR_tpr <- ggplot(tpr_edgeR_merged, aes(x=Cnfnd, y=TPR, group=Method, color=Method)) +
  facet_grid(~Disp) +
  geom_line() +
  #labs(title="edgeR") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_edgeR <- ggarrange(plotlist=list(p_edgeR_fpr, p_edgeR_tpr), nrow=2, 
                     common.legend=TRUE, legend="bottom")#, heights=c(5.5,6))

## DESeq2 only
p_DESeq2_fpr <- ggplot(fpr_DESeq2_merged, aes(x=Cnfnd, y=FPR, group=Method, color=Method)) +
  facet_grid(~Disp) +
  geom_line() +
  geom_hline(yintercept=0.05, col="red", linetype="dashed") +
  labs(title="DESeq2") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
  #theme(axis.text.x=element_blank(), axis.title.x=element_blank()) 
p_DESeq2_tpr <- ggplot(tpr_DESeq2_merged, aes(x=Cnfnd, y=TPR, group=Method, color=Method)) +
  facet_grid(~Disp) +
  geom_line() +
  #labs(title="DESeq2") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_DESeq2 <- ggarrange(plotlist=list(p_DESeq2_fpr, p_DESeq2_tpr), nrow=2, 
                      common.legend=TRUE, legend="bottom")#, heights=c(5.5,6))

#p <- ggarrange(plotlist=list(p_all, p_edgeR, p_DESeq2), nrow=3)
p <- ggarrange(plotlist=list(p_edgeR, p_DESeq2), nrow=2)
```

```{r, echo=FALSE, fig.width=5.5, fig.height=7}
p
```


#### Scatter plots

```{r, echo=FALSE}
if(identical(fpr_merged[,-1], tpr_merged[,-1])){
  perfstats_all <- data.frame(TPR=tpr_merged$TPR, FPR=fpr_merged$FPR, tpr_merged[,-1])
}else{stop("ERROR in merging data frames! - All")}
#identical(perfstats_all[,-1],fpr_merged)

if(identical(fpr_edgeR_merged[,-1], tpr_edgeR_merged[,-1])){
  perfstats_edgeR <- data.frame(TPR=tpr_edgeR_merged$TPR, FPR=fpr_edgeR_merged$FPR, tpr_edgeR_merged[,-1])
}else{stop("ERROR in merging data frames! - edgeR")}
#identical(perfstats_edgeR[,-1], fpr_edgeR_merged)

if(identical(fpr_DESeq2_merged[,-1], tpr_DESeq2_merged[,-1])){
  perfstats_DESeq2 <- data.frame(TPR=tpr_DESeq2_merged$TPR, FPR=fpr_DESeq2_merged$FPR, tpr_DESeq2_merged[,-1])
}else{stop("ERROR in merging data frames! - DESeq2")}
#identical(perfstats_DESeq2[,-1], fpr_DESeq2_merged)

```

```{r, echo=FALSE}
p2_all <- ggplot(perfstats_all, aes(x=FPR, y=TPR, group=Method, color=Method, shape=Method)) +
  geom_point(size=3) +
  facet_grid(Cnfnd~Disp) +
  scale_shape_manual(values=1:nlevels(perfstats_all$Method)) +
  geom_vline(xintercept=0.05, linetype="dashed") +
  labs(title="All methods") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, hjust=1)) 

p2_edgeR <- ggplot(perfstats_edgeR, aes(x=FPR, y=TPR, group=Method, color=Method, shape=Method)) +
  geom_point(size=3) +
  facet_grid(Cnfnd~Disp) +
  scale_shape_manual(values=1:nlevels(perfstats_edgeR$Method)) +
  geom_vline(xintercept=0.05, linetype="dashed") +
  labs(title="edgeR") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, hjust=1)) 

p2_DESeq2 <- ggplot(perfstats_DESeq2, aes(x=FPR, y=TPR, group=Method, color=Method, shape=Method)) +
  geom_point(size=3) +
  facet_grid(Cnfnd~Disp) +
  scale_shape_manual(values=1:nlevels(perfstats_DESeq2$Method)) +
  geom_vline(xintercept=0.05, linetype="dashed") +
  labs(title="DESeq2") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, hjust=1)) 
  
p2 <- ggarrange(p2_edgeR, p2_DESeq2, nrow=2)
```

```{r, echo=FALSE, fig.width=5, fig.height=7}
p2
```



### TPR (sensitivity) vs FDR (1-precision) {.tabset}

```{r, echo=FALSE, results='hide'}
rm(list=ls())
results_dir <- "~/Google Drive/ComBat_seq/DE_disp_confounding_base"
#results_dir <- "~/yuqingz/ComBat_seq/DE_disp_confounding_base"
N_samples <- 20

disp_level_vec <- 1:5
confounding_level_vec <- seq(from=0.5,to=0.2,by=-0.1)
sel_method_list <- list(edgeR=c("BaseIndi.edgeR", #"BaseQuant.edgeR", 
                                "Batch.edgeR", "OneStep.edgeR", "ComBat.lm", 
                                "RUVseq.edgeR", "SVAseq.edgeR", "ComBatseq.edgeR"),
                        DESeq2=c("BaseIndi.DESeq2", #"BaseQuant.DESeq2", 
                                 "Batch.DESeq2", "OneStep.DESeq2", "ComBat.lm", 
                                 "RUVseq.DESeq2", "SVAseq.DESeq2", "ComBatseq.DESeq2"))
```

```{r, echo=FALSE, results='hide'}
fdr_all_lst <- tpr_all_lst <- list()
for(i in 1:length(confounding_level_vec)){
  fdr_all_lst[[i]] <- tpr_all_lst[[i]] <- list()
  for(j in 1:length(disp_level_vec)){
    exp_name <- paste0("simDCbase_N", N_samples, "_dispFC", disp_level_vec[j],
                       "_percent", gsub(".","", confounding_level_vec[i], fixed=T))
    tpr_res <- read.csv(file.path(results_dir, sprintf('tprADJ_%s.csv', exp_name)))
    prec_res <- read.csv(file.path(results_dir, sprintf('precADJ_%s.csv', exp_name)))
    
    tpr_res$BaseQuant.edgeR <- NULL; tpr_res$BaseQuant.DESeq2 <- NULL
    prec_res$BaseQuant.edgeR <- NULL; prec_res$BaseQuant.DESeq2 <- NULL
    
    tpr_all_lst[[i]][[j]] <- as.list(colMeans(tpr_res, na.rm=TRUE))
    fdr_all_lst[[i]][[j]]  <- as.list(colMeans(1-prec_res, na.rm=TRUE))
  }
  names(tpr_all_lst[[i]]) <- names(fdr_all_lst[[i]]) <- paste0("Disp", disp_level_vec)
}
names(tpr_all_lst) <- names(fdr_all_lst) <- paste0("Cnfnd", confounding_level_vec)

## All methods
tpr_merged <- melt(tpr_all_lst); colnames(tpr_merged) <- c("TPR", "Method", "Disp", "Cnfnd")
fdr_merged <- melt(fdr_all_lst); colnames(fdr_merged) <- c("FDR", "Method", "Disp", "Cnfnd")
tpr_merged$Method <- factor(tpr_merged$Method, levels=colnames(tpr_res))
fdr_merged$Method <- factor(fdr_merged$Method, levels=colnames(prec_res))
tpr_merged$Cnfnd <- factor(tpr_merged$Cnfnd, levels=paste0("Cnfnd", confounding_level_vec))
fdr_merged$Cnfnd <- factor(fdr_merged$Cnfnd, levels=paste0("Cnfnd", confounding_level_vec))


## edgeR only
tpr_edgeR_lst <- lapply(tpr_all_lst, function(t1_lst){lapply(t1_lst, function(t2_lst){t2_lst[sel_method_list$edgeR]})})
fdr_edgeR_lst <- lapply(fdr_all_lst, function(f1_lst){lapply(f1_lst, function(f2_lst){f2_lst[sel_method_list$edgeR]})})

tpr_edgeR_merged <- melt(tpr_edgeR_lst); colnames(tpr_edgeR_merged) <- c("TPR", "Method", "Disp", "Cnfnd")
fdr_edgeR_merged <- melt(fdr_edgeR_lst); colnames(fdr_edgeR_merged) <- c("FDR", "Method", "Disp", "Cnfnd")
tpr_edgeR_merged$Method <- factor(tpr_edgeR_merged$Method, levels=sel_method_list$edgeR)
fdr_edgeR_merged$Method <- factor(fdr_edgeR_merged$Method, levels=sel_method_list$edgeR)
tpr_edgeR_merged$Cnfnd <- factor(tpr_edgeR_merged$Cnfnd, levels=paste0("Cnfnd", confounding_level_vec))
fdr_edgeR_merged$Cnfnd <- factor(fdr_edgeR_merged$Cnfnd, levels=paste0("Cnfnd", confounding_level_vec))


## DESeq2 only
tpr_DESeq2_lst <- lapply(tpr_all_lst, function(t1_lst){lapply(t1_lst, function(t2_lst){t2_lst[sel_method_list$DESeq2]})})
fdr_DESeq2_lst <- lapply(fdr_all_lst, function(f1_lst){lapply(f1_lst, function(f2_lst){f2_lst[sel_method_list$DESeq2]})})

tpr_DESeq2_merged <- melt(tpr_DESeq2_lst); colnames(tpr_DESeq2_merged) <- c("TPR", "Method", "Disp", "Cnfnd")
fdr_DESeq2_merged <- melt(fdr_DESeq2_lst); colnames(fdr_DESeq2_merged) <- c("FDR", "Method", "Disp", "Cnfnd")
tpr_DESeq2_merged$Method <- factor(tpr_DESeq2_merged$Method, levels=sel_method_list$DESeq2)
fdr_DESeq2_merged$Method <- factor(fdr_DESeq2_merged$Method, levels=sel_method_list$DESeq2)
tpr_DESeq2_merged$Cnfnd <- factor(tpr_DESeq2_merged$Cnfnd, levels=paste0("Cnfnd", confounding_level_vec))
fdr_DESeq2_merged$Cnfnd <- factor(fdr_DESeq2_merged$Cnfnd, levels=paste0("Cnfnd", confounding_level_vec))
```


#### Line plots {.tabset}

##### Facet by confounding levels

```{r, echo=FALSE, results="hide"}
## All methods
p_fdr <- ggplot(fdr_merged, aes(x=Disp, y=FDR, group=Method, color=Method)) +
  facet_grid(~Cnfnd) +
  geom_line() +
  geom_hline(yintercept=0.05, col="red", linetype="dashed") +
  labs(title="All", y="1 - Precision (FDR)") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_tpr <- ggplot(tpr_merged, aes(x=Disp, y=TPR, group=Method, color=Method)) +
  facet_grid(~Cnfnd) +
  geom_line() +
  labs(y="Sensitivity (TPR)") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_all <- ggarrange(plotlist=list(p_fdr, p_tpr), nrow=2, 
                   common.legend=TRUE, legend="bottom")

## edgeR only
p_edgeR_fdr <- ggplot(fdr_edgeR_merged, aes(x=Disp, y=FDR, group=Method, color=Method)) +
  facet_grid(~Cnfnd) +
  geom_line() +
  geom_hline(yintercept=0.05, col="red", linetype="dashed") +
  labs(title="edgeR", y="1 - Precision (FDR)") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_edgeR_tpr <- ggplot(tpr_edgeR_merged, aes(x=Disp, y=TPR, group=Method, color=Method)) +
  facet_grid(~Cnfnd) +
  geom_line() +
  labs(y="Sensitivity (TPR)") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_edgeR <- ggarrange(plotlist=list(p_edgeR_fdr, p_edgeR_tpr), nrow=2, 
                     common.legend=TRUE, legend="bottom")

## DESeq2 only
p_DESeq2_fdr <- ggplot(fdr_DESeq2_merged, aes(x=Disp, y=FDR, group=Method, color=Method)) +
  facet_grid(~Cnfnd) +
  geom_line() +
  geom_hline(yintercept=0.05, col="red", linetype="dashed") +
  labs(title="DESeq2", y="1 - Precision (FDR)") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_DESeq2_tpr <- ggplot(tpr_DESeq2_merged, aes(x=Disp, y=TPR, group=Method, color=Method)) +
  facet_grid(~Cnfnd) +
  geom_line() +
  labs(y="Sensitivity (TPR)") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_DESeq2 <- ggarrange(plotlist=list(p_DESeq2_fdr, p_DESeq2_tpr), nrow=2, 
                      common.legend=TRUE, legend="bottom")

p <- ggarrange(plotlist=list(p_edgeR, p_DESeq2), nrow=2)
```

```{r, echo=FALSE, fig.width=5, fig.height=7}
p
```


##### Facet by dispersion levels

```{r, echo=FALSE}
## All methods
p_fdr <- ggplot(fdr_merged, aes(x=Cnfnd, y=FDR, group=Method, color=Method)) +
  facet_grid(~Disp) +
  geom_line() +
  geom_hline(yintercept=0.05, col="red", linetype="dashed") +
  labs(title="All", y="1 - Precision (FDR)") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_tpr <- ggplot(tpr_merged, aes(x=Cnfnd, y=TPR, group=Method, color=Method)) +
  facet_grid(~Disp) +
  geom_line() +
  labs(y="Sensitivity (TPR)") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_all <- ggarrange(plotlist=list(p_fdr, p_tpr), nrow=2, 
                   common.legend=TRUE, legend="bottom")

## edgeR only
p_edgeR_fdr <- ggplot(fdr_edgeR_merged, aes(x=Cnfnd, y=FDR, group=Method, color=Method)) +
  facet_grid(~Disp) +
  geom_line() +
  geom_hline(yintercept=0.05, col="red", linetype="dashed") +
  labs(title="edgeR", y="1 - Precision (FDR)") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_edgeR_tpr <- ggplot(tpr_edgeR_merged, aes(x=Cnfnd, y=TPR, group=Method, color=Method)) +
  facet_grid(~Disp) +
  geom_line() +
  labs(y="Sensitivity (TPR)") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_edgeR <- ggarrange(plotlist=list(p_edgeR_fdr, p_edgeR_tpr), nrow=2, 
                     common.legend=TRUE, legend="bottom")

## DESeq2 only
p_DESeq2_fdr <- ggplot(fdr_DESeq2_merged, aes(x=Cnfnd, y=FDR, group=Method, color=Method)) +
  facet_grid(~Disp) +
  geom_line() +
  geom_hline(yintercept=0.05, col="red", linetype="dashed") +
  labs(title="DESeq2", y="1 - Precision (FDR)") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_DESeq2_tpr <- ggplot(tpr_DESeq2_merged, aes(x=Cnfnd, y=TPR, group=Method, color=Method)) +
  facet_grid(~Disp) +
  geom_line() +
  labs(y="Sensitivity (TPR)") +
  theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) 
p_DESeq2 <- ggarrange(plotlist=list(p_DESeq2_fdr, p_DESeq2_tpr), nrow=2, 
                      common.legend=TRUE, legend="bottom")

p <- ggarrange(plotlist=list(p_edgeR, p_DESeq2), nrow=2)
```

```{r, echo=FALSE, fig.width=5.5, fig.height=7}
p
```


#### Scatter plots

Similar plot (sensitivity vs 1-FDR) as Figure 6 in DESeq2 paper.

```{r, echo=FALSE}
if(identical(fdr_merged[,-1], tpr_merged[,-1])){
  perfstats_all <- data.frame(TPR=tpr_merged$TPR, FDR=fdr_merged$FDR, tpr_merged[,-1])
}else{stop("ERROR in merging data frames! - All")}
#identical(perfstats_all[,-1],fdr_merged)

if(identical(fdr_edgeR_merged[,-1], tpr_edgeR_merged[,-1])){
  perfstats_edgeR <- data.frame(TPR=tpr_edgeR_merged$TPR, FDR=fdr_edgeR_merged$FDR, tpr_edgeR_merged[,-1])
}else{stop("ERROR in merging data frames! - edgeR")}
#identical(perfstats_edgeR[,-1], fdr_edgeR_merged)

if(identical(fdr_DESeq2_merged[,-1], tpr_DESeq2_merged[,-1])){
  perfstats_DESeq2 <- data.frame(TPR=tpr_DESeq2_merged$TPR, FDR=fdr_DESeq2_merged$FDR, tpr_DESeq2_merged[,-1])
}else{stop("ERROR in merging data frames! - DESeq2")}
#identical(perfstats_DESeq2[,-1], fdr_DESeq2_merged)

```

```{r, echo=FALSE}
p2_all <- ggplot(perfstats_all, aes(x=FDR, y=TPR, group=Method, color=Method, shape=Method)) +
  geom_point(size=3) +
  facet_grid(Cnfnd~Disp) +
  scale_shape_manual(values=1:nlevels(perfstats_all$Method)) +
  geom_vline(xintercept=0.05, linetype="dashed") +
  labs(title="All methods", x="1 - Precision (FDR)", y="Sensitivity (TPR)") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, hjust=1)) 

p2_edgeR <- ggplot(perfstats_edgeR, aes(x=FDR, y=TPR, group=Method, color=Method, shape=Method)) +
  geom_point(size=3) +
  facet_grid(Cnfnd~Disp) +
  scale_shape_manual(values=1:nlevels(perfstats_edgeR$Method)) +
  geom_vline(xintercept=0.05, linetype="dashed") +
  labs(title="edgeR", x="1 - Precision (FDR)", y="Sensitivity (TPR)") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, hjust=1)) 

p2_DESeq2 <- ggplot(perfstats_DESeq2, aes(x=FDR, y=TPR, group=Method, color=Method, shape=Method)) +
  geom_point(size=3) +
  facet_grid(Cnfnd~Disp) +
  scale_shape_manual(values=1:nlevels(perfstats_DESeq2$Method)) +
  geom_vline(xintercept=0.05, linetype="dashed") +
  labs(title="DESeq2", x="1 - Precision (FDR)", y="Sensitivity (TPR)") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, hjust=1)) 
  
p2 <- ggarrange(p2_edgeR, p2_DESeq2, nrow=2)
```

```{r, echo=FALSE, fig.width=5, fig.height=7}
p2
```



## Thoughts & future work

* 
